Effects of bariatric surger and weight loss on resting-state functional connectivity

### 2020-12-03 manuscript/draft_rsfmri.Rmd
Moved files"

```
\newpage
# Abstract

\newpage
# Introduction
<!--Suggested Order of Introduction
1. Obesity is a major health concern. [... some more details about metabolic and mortality risk, especially related to morbid obesity ...] More and more recognized, midlie obesity also impacts later cognitive decline and dementia risk.
Obesity has been associated with differences in brain structure and function.
Regarding structural alterations, longitudinal studies have indicated that higher body mass index is related to increased cortial thinning in midlife (path through life study), potentially mediated by inflammation or glucose dysregulation... Similarly, white matter microstructure and may be related to adverse metabolic effects (lampe)
Differences in brain function in obesity have been reported in the reward network during task-based fMRI, highlighting the role of reward processing in obesity development and maintainance.
Resting state fMRI studies have also shown differences in connectivity in diverse brain networks, including reward and salience network, visual networks and the default mode network. In part, these differences have been related to metabolic and behavioral differences, yet there may be an important confounding due to head motion.
In this study, we aim to investigate the impact of bariatric surery on..
-->

<!-- obesity and interventions -->
Over the past decades, the global burden of obesity has increasingly been confronted as a matter of urgency by research.
_something about comorbidities_
Besides metabolic dysfunctions, obesity has been consistently associated with reduced gray matter volume, less intact white matter structure and reduced cognitive function [@Beyer_2019; @Zhang_2018]. 
Luckily, dietary interventions such as caloric restriction and related metabolic improvements, and weight loss have been suggested to improve neuronal plasticity and cognitive performance [@Witte_2009; @Zhang_2018; @Veronese_2017].
In some cases behavioral intervention fail to yield the desired success regarding weight loss. There are non-behavioral alternatives at hand.
<!-- bars most invasive (brain metabolism), last resort -->
Bariatric surgery is one promising option to combat morbid obesity, as it rapidly improves weight status, metabolic dysfunctions, and co-morbidities, such as diabetes [@]. Roux-en-Y gastric bypass (RYGB) is the most common procedure as it not only restricts the food intake by reducing stomach volume but also affects the absorption of macronutrients and the hormonal profile [@]. <!-- ref for "common", ref for procedure -->
Other widely used procedures are vertical sleeve gastrectomy (VSG) and gastric banding (GB) which only restrict food consumption. <!-- mehr zu Veränderung in Gasto-intestinaler Physiologie -->

<!-- Neural mechanisms/concurrent plasticity? -->
Precise mechanisms are yet to be elucidated. As a first step, a comprehensive understanding of the concurrent physiological changes underlying weight loss and recovery is needed. This will inform inferences about potential mechanisms and either support interventions proven to be effective or ideally allow imitating them will less invasive alternatives.

Recently, research focuses have slowly shifted to investigate brain metabolism structural and functional changes after surgery.
<!-- severe nutritional stress [@Crujeiras_2017] -->

<!-- Reward network -->
Little is known about the potential beneficial effects of bariatric surgery on brain function in the obese population, in particular with regard to hedonic motivations for food, which play a fundamental role in weight relapse.
<!-- mechanism reward and appetite, overating -->
Besides malabsorption and restricted food ingestion, patients report post-surgical changes in food preferences and eating behavior. Those psychophysiological and behavioral effects may be related to post-surgical changes in visceral sensory perceptions and endocrinological signaling. Corresponding information is processed by the fronto-mesolimbic system and affects appetite and reward signalling which may influence weight regulation [@Mulla_2017, @Nudel_2019].
<!-- results in lean -->
The reward network entailing the ventromedial prefrontal cortex (vmPFC), the striatum (STR) but also the anterior insula (antIns) has been linked to food valuation processes in decision-making in lean individuals [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018; @Wiemerslage_2016].
<!-- results in obese longitudinal -->
Similarly, several studies have found differences in resting-state activity in reward regions depending on weight status.

Obese individuals exhibited increased connectivity between frontal and mesolimbic regions and weaker connectivity within regions of the mesolimbic system (encompassing limbic and striatal regions). These differences seemed to normalize, as trends reversed when looking at the resting state of obese individuals after compared to before the surgery [@Duan_2020].

On the other hand, BMI prior to the surgery was associated with changes in connectivity [@Duan_2020].

s concerning such processes following a surgical intervention are inconclusive.

<!-- important differences in the response to visual food-related cues between thin and reduced-obese individuals that place the latter group at an increased risk of  -->

<!-- Though there exists strong evidence for important differences in the .. between lean or normal-weight individuals and obese individuals, few studies have investigated the ... and even fewer compared reduced-obese individuals reduced-obese individuals -->
<!-- [@Garcia_Garcia_2012] -->

<!-- Default mode network --> 
Weight regulation and energy expenditure is dependent on a complex interplay of external and internal sensory processing, as well as molecular homeostasis regulation. Obesity may be cause or effect of a decompensation of the psychophysiological system that ensures homeostasis.
The default mode network is commonly associated with processing of self-referent as well as introspective information, such as hunger sensation [@;@] and appetite [@Tregellas_2011]. <!-- parietal activation correlated with appetite measures after an overnight fast [@Tregellas_2011]. --> It is mainly composed of the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (MPFC) and inferior and lateral parietal cortex (IPC, LPC; approx. BA 39) [@Raichle_2015]. The right ventral anterior cingulate cortex (VACC), superior frontal cortex (SFC), and inferior temporal cortex (ITC) show weaker but also reliable association with the PCC over the course of one year. Even broader conceptions count left VACC, middle temporal gyrus (MTG), and parahippocampal cortex, although associations with the PCC are not as stable [@Chou_2012].
<!-- difference lean obese --> 
Functional connectivity of this network has been shown to be linked to BMI [@Beyer_2017] and altered after bariatric surgery [@Olivo_2016]. Generally, studies show an increased activation in the default mode network, especially in the precuneus and posterior cingulate cortex (PCC) in obese individuals [@Kullmann_2011; @].
A longitudinal study that confirmed the stronger connectivity (of PCC with anterior STG and anterior temporal fusiform gyrus and the Precuneus with the anterior STG). However, default mode connectivity in these regions seemed weakened over the course of one year. Because differences between lean controls and obese individuals that existed prebariatric surgery were no longer detectable 1 year after the surgery, the authors argue for a normalization in functional connectivity<!-- and PCC/PC connectivity with anterior STG are most affected by the surgery -->[@Olivo_2017].
This is consistent with other findings of reduction of functional connectivity of these and other regions of the default mode network, such as the VMPFC [@Li_2018] <!-- with the dorsolateral prefrontal cortex (DLPFC) --> after the surgery.

The exact mechanisms of how bariatric surgery affects functional plasticity needs to be investigated. Among patients undergoing surgery, change in BMI was negatively correlated with the changes in connectivity of the default mode network, namely between the PCC/precuneus and the DLPFC [@Li_2018].

On the other hand, connectivity also served as a predictor of weight loss after surgery.

On the other hand, there are also indicators of reduced functional integrity, e.g. for the anterior cingulate cortex (ACC) at resting state compared to lean individuals [@Kullmann_2011].

<!-- novelty statement --> As described previously, several studies have investigated the effects of bariatric surgery on the intrinsic functional connectivity in patients. Yet, to our knowledge there is no quasi-experimental study, comparing a cohort of patients to a cohort with similar characteristics but unexposed to the intervention, i.e. bariatric surgery, over an extended time period as would be demanded to draw more substantiated inference [@George_2016].
<!-- hypotheses --> In this study, we aim to bridge this gap and investigate possible dynamics of within-reward network connectivity before and after bariatric surgery by comparison with a contemporaneously measured waiting-control group. Moreover, we will explore whether the changes in functional connectivity are associated with the amount of weight lost after bariatric surgery treatment in obesity to explore a potential gradient link. In addition, we will investigate potential changes in within-network functional connectivity after surgery in other resting-state networks such as default mode network (DMN).
Participants are measured at baseline, at 6 months and at 24 months to capture both phases of rapid weight-loss and maintenance as expected from non-surgical interventions [@Shai_2008].

<!-- group effect because direct effect of intervention surgery -> rsfmri (cares lessfreq, less preoccupation with hunger, less/more reward sensitivity) vs. sugery -> BMI -> rsfmri (neural signalling, fatty tissue) -->

@@

<!-- DESCRIPTIVE STATISTICS --> The final sample entailed morbidly obese individuals (n = `r nrow(subj_df)`, `r nrow(filter(subj_df,Sex == "f"))` female; ages `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`), that either underwent surgery (n = `r cs["Sum","KG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female). The study design was quasi-experimental, as patient in the waiting control list had also been recommended to undergo bariatric surgery, but were still waiting for their health insurance's approval.
```{r test wd}
print(getwd())
```

```{r baseline characteristics, include=FALSE}
dfBL <- save_txt_for_swe(group = "all", tp = "BL")

# distributions per group
# BMI 
p <- ggplot(data=dfBL, aes(group = condition, x=BMI, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()
# log mean FD
p <- ggplot(data=dfBL, aes(group = condition, x=logmFD, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E55", "#C0004555"),
    labels = c("intervention", "control")) +
  theme_bw()
# Age
p <- ggplot(data=dfBL, aes(group = condition, x=Age_BL, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()

tab_age <- psych::describeBy(data=dfBL, x=dfBL$Age_BL, group=dfBL$condition, mat=T)

# Sex
p <- ggplot(data = dfBL, aes(x = condition, fill = Sex)) + 
  geom_bar(position = "dodge") +
  scale_fill_manual(labels = c("female", "male"),
                    values = c("#00305E55", "#C0004555")
  )
tab_sex <- table(dfBL$condition, dfBL$Sex)
```
Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and log mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). 
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
```{r tabdescr, echo=FALSE}

>>>>

```
# Abstract

# Introduction

@@

<!-- DESCRIPTIVE STATISTICS --> The final sample entailed morbidly obese individuals (n = `r nrow(subj_df)`, `r nrow(filter(subj_df,Sex == "f"))` female; ages `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`), that either underwent surgery (n = `r cs["Sum","KG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female). The study design was quasi-experimental, as patient in the waiting control list had also been recommended to undergo bariatric surgery, but were still waiting for their health insurance's approval.
```{r baseline characteristics, include=FALSE}
dfBL <- save_txt_for_swe(group = "all", tp = "BL")

# distributions per group
# BMI 
p <- ggplot(data=dfBL, aes(group = condition, x=BMI, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()
# log mean FD
p <- ggplot(data=dfBL, aes(group = condition, x=logmFD, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E55", "#C0004555"),
    labels = c("intervention", "control")) +
  theme_bw()
# Age
p <- ggplot(data=dfBL, aes(group = condition, x=Age_BL, fill=condition)) +
  geom_density(alpha=0.6) + 
  scale_fill_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")) +
  theme_bw()
tab_age <- psych::describeBy(data=dfBL, x=dfBL$Age_BL, group=dfBL$condition, mat=T)
# Sex
p <- ggplot(data = dfBL, aes(x = condition, fill = Sex)) + 
  geom_bar(position = "dodge") +
  scale_fill_manual(labels = c("female", "male"),
                    values = c("#00305E55", "#C0004555")
  )
tab_sex <- table(dfBL$condition, dfBL$Sex)
```
Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and log mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
```{r tabdescr, echo=FALSE}
